Solving the maximum independant set problem by generic programming einsum networks \thanks{\funding{...}}

### 2021-08-09 paper/paper.tex
upate

\documentclass[review,onefignum,onetabnum]{siamart190516}
%\usepackage{natbib}
%\usepackage[sort]{cite}
\pdfoutput=1
%\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usetikzlibrary{snakes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithmic}
\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

\begin{abstract}
	Solving the maximum independent set size problem by mapping the graph to an einsum network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
\end{abstract}

@@

\section{Introduction}
In this work, we introduce a new method to solve the classic graph problem, finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
Finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbituary small possitive $\epsilon$.
The naive algorithm of enumerating all configuration space gives a $2^{|V|}$ time solution.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm gives a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving this problem better not only because it is a NP-complete problem that directly related to other NP-complete prolems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
In these applications, knowing the MIS size and one of the optimal solution is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all indenepent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements.

\section{Einsum network}
The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of multiplication between two matrices to the contraction between multiple tensors.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
It is denoted as $C_i^k = A_i^j B_j^k$ in the Einstein's original notation, where the paired subscript and superscript $j$ is a dummy index summed over.
One can map a tensor network to a muti-graph with open edges by viewing a tensor in the expression on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^k B^l_k V^j$, where its graphical notation as the following.
One can easily check a label in a tensor network representation appears precisely twice.

@@

Numpy programmers make a generalization of this notation by not restricting the number of times a label is used by tensors,
hence whether an index appears as a superscript or a subscript makes no sense now.
It has different names in different context, like sum-product network and factor graph~\cite{Bishop2006}.
The graphical representation of an einsum is a hypergraph, where an edge (degree of freedom) can be shared by a arbituary number of nodes (tensors).

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is an einsum but not a tensor network, it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Its hypergraph representation is as the following, where we use different color to annotate different hyperedges.

@@

\section{Independence polynomial}
One can encode the independent set problem on graph $G$ to an einsum network by placing a rank one tensor of size $2$ on vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
$x_i$ is a variable.
The contraction of such an einsum network gives
\begin{equation}
    A(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output.
Let $x_i = x$, then the product over vertex tensors gives a factor $x^k$, where $k=\sum_i s_i$ is the vertex set size, 
and the product over edge tensors gives a factor $0$ for configurations not being an independent set.
The contraction of this einsum network gives the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
By mapping the independence polynomial solving problem to the einsum network contraction, one can take the advantage of recently developed techiniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$, where $tw(G)$ is the treewidth of the line graph of a tensor network, here it corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize fast basic linear algebra subprograms (BLAS) functions for certain tensor element types.

@@

        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}
    \begin{align*}
        \sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}.
    \end{align*}
\end{example}

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relation must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

In the rest of this paper, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the einsum network generic~\cite{Stepanov2014}.

@@

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of $m+1$ random real numbers, e.g. $\{0, 1, 2, \ldots, m\}$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is esentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

@@

\subsection{The finite field algebra approach}
It sounds a bit over ambitious to compute the independence polynomial regorously using integer number types only,
because the fixed width integer types are often too small to store the countings,
while big integer with arbituary precision can be very slow and imcompatible with graphic processing units (GPU) devices.
The solution we found is to computation on a finite field algebra $GF(p)$

\begin{align}
\begin{split}
    x ~\oplus~ y &= x+y\pmod p,\\
    x ~\odot~ y &= xy\pmod p,\\
    \mymathbb{0} &= 0,\\
    \mymathbb{1} &= 1.
\end{split}
\end{align}

@@

\section{Computing maximum independent set size and its corresponding degeneracy and configurations}
Obtaining the maximum independent set size and its degeneracy can be computational more efficient. Let $x=\infty$, then the independence polynomial becomes
\begin{equation}
I(G, \infty) = a_k \infty^{\alpha(G)},
\end{equation}
where the lower orders terms disappear automatically. We can define a new algebra as
\begin{align}
\begin{split}
    a_x\infty^x \oplus a_y\infty^y &= \begin{cases}
        (a_x + a_y)\infty^{\max(x,y)}, & x = y\\
        a_y\infty^{\max(x,y)}, & x < y\\
        a_x\infty^{\max(x,y)}, & x > y
    \end{cases}\\
    a_x\infty^x \odot a_y\infty^y &= a_x a_y\infty^{x+y}\\
    \mymathbb{0} &= 0\infty^{-\infty}\\
    \mymathbb{1} &= 1\infty^{0}
\end{split}
\end{align}
In the program, we only store the power $x$ and the corresponding factor $a_x$ that initialized to $1$.
This algebra is consistent with the one we derived in ~\cite{Liu2021} that uses the tropical tensor network for solving spin glass ground states.
If one is only interested in obtaining $\alpha(G)$, he can drop the factor parts, then the algebra of $x$ becomes the max-plus tropical algebra~\cite{Maclagan2015,Moore2011}.

>>>>

\documentclass[review,onefignum,onetabnum]{siamart190516}
%\usepackage{natbib}
%\usepackage[sort]{cite}
\pdfoutput=1
%\usepackage[colorlinks=true,urlcolor=blue,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amssymb}
\usepackage{tabularx}
\usepackage{quoting}
\usepackage{upquote}
\usepackage{subcaption}
\usepackage{multicol}
\usepackage{cancel}
\usepackage[framemethod=TikZ]{mdframed}
\usetikzlibrary{shapes}
\usetikzlibrary{snakes}
\usepackage{wrapfig}
%\usepackage{caption}
%\usepackage[plain]{algorithmic}
\usepackage[linesnumbered, ruled, vlined, algo2e]{algorithm2e}
\usepackage{algpseudocode}
\usepackage{rotating}
%\usepackage{cite}
\usepackage{booktabs}
%\usepackage{unicode-math}
%\usepackage{algorithm}% http://ctan.org/pkg/algorithm
%\usepackage{algpseudocode}% http://ctan.org/pkg/algpseudocode
\usepackage{xcolor}% http://ctan.org/pkg/xcolor
\makeatletter
\newsavebox{\@brx}
\newcommand{\llangle}[1][]{\savebox{\@brx}{\(\m@th{#1\langle}\)}%
  \mathopen{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\newcommand{\rrangle}[1][]{\savebox{\@brx}{\(\m@th{#1\rangle}\)}%
  \mathclose{\copy\@brx\kern-0.5\wd\@brx\usebox{\@brx}}}
\makeatother
\usepackage{bbm}
\usepackage{jlcode}
\usepackage{graphicx}
\usepackage{amsmath,color}
\usepackage{mathrsfs}
\usepackage{float}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{indentfirst}
\usepackage{txfonts}
\usepackage[epsilon, tsrm, altpo]{backnaur}

@@

\begin{abstract}
	Solving the maximum independent set size problem by mapping the graph to an einsum network. 
    We show how to obtain the maximum independent set size, the independence polynomial and optimal configurations of a graph by engineering the tensor element algebra.
    We also show how to analyse the local properties of a graph by contracting an open einsum network.
\end{abstract}

@@

\section{Introduction}
In this work, we introduce a tensor based framework to study the famous graph problem of finding independent sets.
Given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
Finding the maximum independent set (MIS) size $\alpha(G) \equiv \max_{I}|I|$ belongs to the complexity class NP-complete~\cite{Hastad1996}, which is unlikely to be decided in polynomial time.
It is hard to even approximate this size in polynomial time within a factor $|V|^{1-\epsilon}$ for an arbituarily small possitive $\epsilon$.
The naive algorithm of enumerating all configuration space gives a $2^{|V|}$ time solution.
More efficient algorithms to compute the MIS size exactly includes the branching algorithm and dynamic programming.
Without changing the fact of exponential scaling in computing time, the branching algorithm gives a smaller base.
For example, in ~\cite{Xiao2017}, a sophisticated branching algorithm gives a time complexity $1.1893^n n ^{O(1)}$.
The dynamic programming approach~\cite{Courcelle1990,Fomin2013} works better for graphs with small tree width $tw(G)$, it gives an algorithms of complexity $O(2^{tw(G)}tw(G)n)$.
People are interested in solving the independent set problem better not only because it is a NP-complete problem that directly related to other NP-complete prolems like maximal cliques and vertex cover~\cite{Moore2011},
but also for its close relation with physical applications like hard spheres lattice gas model~\cite{Dyre2016}, and Rydberg hamiltonian~\cite{Pichler2018}.
However, in these applications, knowing the MIS size and one of the optimal solution is not the only goal.
%A set is independent if and only if it is a clique in the graph's complement, so the two concepts are complementary.
%A set is independent if and only if its complement is a vertex cover. Therefore, the sum of the size of the largest independent set $\alpha (G)$ and the size of a minimum vertex cover $\beta (G)$ is equal to the number of vertices in the graph.
People often ask different questions about independent sets in order to understand the landscape of their models better.
These questions includes but not limited to, counting all independent sets, obtaining all indenepent sets of size $\alpha(G)$ and $\alpha(G)-1$,
counting the number of (maximal) independent sets of different sizes, and understanding the effect of a local gadget.
In this work, we attack this problem by mapping it to an generic ``einsum'' network.
It does not give a better time complexity comparing to dynamic programming, but is versatile enough to answer the above questions by engineering the tensor elements with minimum effort.

\section{Einsum network}
%The word ``einsum'' is a shorthand for Einstein's summation, however, modern einsum notation in program is actually invented by a group of programmers.
Einstein's notation is originally proposed as a generalization to of binary matrix multiplication to n-ary tensor contraction.
Let $A, B$ be two matrices, the matrix multiplication is defined as $C_{ik} = \sum_{j}A_{ij}B_{jk}$.
In Einstein's notation, it is denoted as $C_i^k = A_i^j B_j^k$, where the paired subscript and superscript $j$ is a dummy index summed over, hence each index appears precisely twice.
When we have multiple tensors doing the above sum-product operation, we get a tensor network~\cite{Orus2014}.
A tensor network has a nice a muti-graph with open edges. We view a tensor on the right hand side as a vertex in a graph, a label pairing two tensors as an edge, and the remaining unpaired labels as open edges.

\begin{example}
    A tensor networks $C_i^l = A_{ij}^k B^l_k V^j$ has the following graphical representation.

@@

Einsum network is a generalization of tensor network by not restricting the number of times a label appears,
hence whether an index is a superscript or a subscript makes no sense now.
It is also called sum-product network or factor graph~\cite{Bishop2006} in some contexts.
The graphical representation of an einsum is a hypergraph, where an edge (label) can be shared by an arbituary number of vertices (tensors).

\begin{example}
$C_{ijk} = A_{jkm} B_{mil} V_{jm}$ is an einsum network, it represents $C_{ijk} = \sum_{ml}A_{jkm} B_{mia} V_{jm}$.
Its hypergraph representation is as the following, where we use different color to annotate different hyperedges.

@@

\section{Independence polynomial}
One can encode the independent set problem on graph $G$ to an einsum network by placing a rank one tensor of size $2$ on vertex $i$
\begin{equation}
    W(x_i)_{s_i} = \left(\begin{matrix}
        1 \\
        x_i
    \end{matrix}\right)_{s_i},\label{eq:wtensor}
\end{equation}
and a rank two tensor of size $2 \times 2$ on edge $(i,j)$
\begin{equation}
    B_{s_i s_j} = \left(\begin{matrix}
        1  & 1\\
        1 & 0
    \end{matrix}\right)_{s_is_j},\label{eq:btensor}
\end{equation}
where a tensor index $s_i$ is a boolean variable that being 1 if vertex $i$ is in the independent set, 0 otherwise.
It corresponds to a hyperedge in the hypergraph.
$x_i$ is a variable.
The contraction of such an einsum network gives
\begin{equation}
    A(G, \{x_1,\ldots,x_n\}) = \sum\limits_{s_1, s_2, \ldots, s_n = 0}^{1} \prod\limits_{i=1}^n W(x_i)_{s_i} \prod\limits_{(i,j) \in E(G)} B_{s_i s_j}.
\end{equation}
Here, the einsum runs over all vertex configurations $\{s_1,\ldots,s_n\}$ and accumulates the product of tensor elements to the scalar output.
Let $x_i = x$, then the product over vertex tensors gives a factor $x^k$, where $k=\sum_i s_i$ is the vertex set size, 
and the product over edge tensors gives a factor $0$ for configurations not being an independent set.
The contraction of this einsum network gives the independence polynomial~\cite{Ferrin2014, Harvey2017} of $G$
\begin{equation}
I(G, x) = \sum_{k=1}^{\alpha(G)} a_k x^k,
\end{equation}
where $a_k$ is the number of independent sets of size $k$ in $G$, and $\alpha(G)$ is the maximum independent set size.
The benefit of mapping the independent set problem to the einsum network is one can take the advantage of recently developed techiniques in tensor network based quantum circuit simulations ~\cite{Gray2021,Pan2021},
where people evaluate a tensor network by pairwise contracting tensors in a heuristic order.
A good contraction order can reduce the time complexity significantly, at the cost of having a space overhead of $O(2^{tw(G)})$, where $tw(G)$ is the treewidth of the line graph of a tensor network, 
here it corresponds to the original graph $G$ that we mapped from.~\cite{Markov2008}
The pairwise tensor contraction also makes it possible to utilize fast basic linear algebra subprograms (BLAS) functions for certain tensor element types.

@@

        \def\dx{0};
        \def\r{0.25cm}
        \filldraw[fill=black] (\dx,0) circle [radius=\r];
        \filldraw[fill=black] (\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,0) circle [radius=\r];
        \filldraw[fill=black] (1.5+\dx,1.5) circle [radius=\r];
        \filldraw[fill=black] (2.5+\dx,2.5) circle [radius=\r];
        \draw [black,thick] (\dx,0) -- (\dx,1.5);
        \draw [black,thick] (\dx,0) -- (1.5+\dx,0);
        \draw [black,thick] (\dx,1.5) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (1.5+\dx,1.5);
        \draw [black,thick] (1.5+\dx,0) -- (\dx,1.5);
        \draw [black,thick] (2.5+\dx,2.5) -- (1.5+\dx,1.5);
        \node[color=white] at (\dx,0) {a};
        \node[color=white] at (\dx,1.5) {b};
        \node[color=white] at (1.5+\dx,0) {c};
        \node[color=white] at (1.5+\dx,1.5) {d};
        \node[color=white] at (2.5+\dx,2.5) {e};
        \def\dx{5};
        \def\r{0.25cm}
        \foreach \x/\y/\e in {0.75/0/ac, 0/0.75/ab, 1.5/0.75/cd, 0.75/1.5/bd, 0.75/0.75/bc, 2/2/de}
            \node[color=white,fill=black,dot=2*\r] at (\x+\dx,\y) (\e) {\scriptsize B};
        \foreach \x/\y/\v in {0/0/a, 0/1.5/b, 1.5/0/c, 1.5/1.5/d, 2.5/2.5/e}
            \node[color=black] at (\x+\dx,\y) (\v) {$s_\v$};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=white,fill=black,dot=\r] at (\x+\dx,\y) (\v\v) {};
        \foreach \x/\y/\v in {-0.5/-0.5/a, -0.5/2.0/b, 2.0/-0.5/c, 2.0/1.0/d, 3.0/2.0/e}
            \node[color=black] at (\x+\dx+0.5,\y) {\scriptsize $W(x_\v)$};
        \draw [cyan,thick] (a) -- (aa);
        \draw [cyan,thick] (a) -- (ab);
        \draw [cyan,thick] (a) -- (ac);
        \draw [blue,thick] (b) -- (bb);
        \draw [blue,thick] (b) -- (ab);
        \draw [blue,thick] (b) -- (bc);
        \draw [blue,thick] (b) -- (bd);
        \draw [red,thick] (c) -- (cc);
        \draw [red,thick] (c) -- (ac);
        \draw [red,thick] (c) -- (bc);
        \draw [red,thick] (c) -- (cd);
        \draw [green,thick] (d) -- (dd);
        \draw [green,thick] (d) -- (bd);
        \draw [green,thick] (d) -- (de);
        \draw [green,thick] (d) -- (cd);
        \draw [orange,thick] (e) -- (ee);
        \draw [orange,thick] (e) -- (de);
    \end{tikzpicture}}
    The contraction of this network can be done in a pairwise order.
    \begin{align*}
        &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a)_{s_a} W(x_b)_{s_b} W(x_c)_{s_c} W(x_d)_{s_d} W(x_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}.\\
        =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e)_{s_e}\right) W(x_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c)_{s_c}\right)\right)\right.\\
        &\phantom{XXX}\left.\left(B_{s_bs_c}\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a)_{s_a}\right)\right)\right)\\
        =&1 + x_a + x_b + x_c + x_d + x_e + x_ax_d + x_ax_e + x_cx_e + x_bx_e
    \end{align*}
\end{example}

Before contracting the einsum network and evaluating the independence polynomial numerically, let us first give up thinking $0$s and $1$s in tensors $W(x)$ and $B$ as regular computer numbers such as integers and floating point numbers.
Instead, we treat them as the additive identity and multiplicative identity of a commutative semiring.
A semiring is a ring without additive inverse, while a commutative semiring is a semiring that multiplication commutative.
To define a commutative semiring with addition algebra $\oplus$ and multiplication algebra $\odot$ on a set $R$, the following relations must hold for arbituary three elements $a, b, c \in R$.
\begin{align*}
(a \oplus b) \oplus c = a \oplus (b \oplus c) & \hspace{5em}\text{$\triangleright$ commutative monoid $\oplus$ with identity $\mymathbb{0}$}\\
a \oplus \mymathbb{0} = \mymathbb{0} \oplus a = a &\\
a \oplus b = b \oplus a &\\
&\\
(a \odot b) \odot c = a \odot (b \odot c)  &   \hspace{5em}\text{$\triangleright$ commutative monoid $\odot$ with identity $\mymathbb{1}$}\\
a \odot  \mymathbb{1} =  \mymathbb{1} \odot a = a &\\
a \odot b = b \odot a &\\
&\\
a \odot (b\oplus c) = a\odot b + a\odot c  &  \hspace{5em}\text{$\triangleright$ left and right distributive}\\
(a\oplus b) \odot c = a\odot c \oplus b\odot c &\\
&\\
a \odot \mymathbb{0} = \mymathbb{0} \odot a = \mymathbb{0}
\end{align*}

In the following, we show how to obtain the independence polynomial, the maximum independent set size and optimal configurations of a general graph $G$ by designing tensor element types as commutative semirings,
i.e. making the einsum network generic~\cite{Stepanov2014}.

@@

\subsection{The fitting and Fourier transformation approaches}
Let $m=\alpha(G)$ be the maximum independent set size and $X$ be a set of real numbers of cardinality $m+1$.
We compute the einsum contraction for each $x_i \in X$ and obtain the following relations
\begin{align}
    \begin{split}
a_0 + a_1 x_1 + a_1 x_1^2 + \ldots + a_m x_1^m &= y_0\\
a_0 + a_1 x_2 + a_2 x_2^2 + \ldots + a_m x_2^m &= y_1\\
\ldots&\\
a_0 + a_1 x_m + a_2 x_m^2 + \ldots + a_m x_m^m& = y_m
    \end{split}
\end{align}
The polynomial fitting between $X$ and $Y = \{y_0, y_1, \ldots, y_m\}$ gives us the factors.
The polynomial fitting is esentially about solving the following linear equation
\begin{align}
\left(\begin{matrix}
1 & x_1 & x_1^2 & \ldots & x_1^m \\
1 & x_2 & x_2^2 & \ldots & x_2^m \\
\vdots & \vdots & \vdots &\ddots & \vdots \\
1 & x_m & x_m^2 & \ldots & x_m^m
\end{matrix}\right)
\left(\begin{matrix}
a_0 \\ a_1 \\ \vdots \\ a_m
\end{matrix}\right)
= \left(\begin{matrix}
y_0 \\ y_1 \\ \vdots \\ y_m
\end{matrix}\right).\label{eq:lineareq}
\end{align}

@@

\subsection{The finite field algebra approach}
It sounds a bit over ambitious to compute the independence polynomial regorously using integer number types only,
because the fixed width integer types are often too small to store the countings,
while big integer with varying width can be very slow and imcompatible with graphic processing units (GPU) devices.
This problem will be solved if we computate on a finite field algebra $GF(p)$

\begin{align}
\begin{split}
    x ~\oplus~ y &= x+y\pmod p,\\
    x ~\odot~ y &= xy\pmod p,\\
    \mymathbb{0} &= 0,\\
    \mymathbb{1} &= 1.
\end{split}\label{eq:finitefield}
\end{align}

@@

\subsection{Maximal independence polynomial}
Let us denote the neighbor of a vertex $v$ as $N(v)$ and $N[v] = N(v)\cup \{v\}$.
A maximal independent set $I_m$ is an independent sets that there does not exist a vertex $v$ that $N[v] \cap I_m = \emptyset$.
Let us modify the einsum network for computing independence polynomial to count maximal independent sets.
We define a tensor on $N[v]$ to capture this property
\begin{align}
    T(x)_{s_1,s_2,\ldots,s_{|N(v)|},s_v} = \begin{cases}
        s_vx & s_1=s_2=\ldots=s_{|N(v)|}=0,\\
        1-s_v& otherwise.\\
    \end{cases}
\end{align}
As an example, for a vertex of degree 2, the resulting rank 3 tensor is
\begin{align}
    T(x)=\left(\begin{matrix}
    \left(\begin{matrix}
        0 &1 \\
        1 &1
    \end{matrix}\right)\\
    \left(\begin{matrix}
        x &0 \\
        0 &0
    \end{matrix}\right)
    \end{matrix}\right).
\end{align}

We do the same computation as independence polynomial, the coefficients of resulting polynomial gives the counting of maximal independent sets, or the maximal independence polynoimal.
The treewidth of this new tensor network is often larger than the one for computing independence polynomial.
However, in many sparse graphs, this tensor network contraction approach is still much faster than computing the maximal cliques on its complement by applying the Bron Kerbosch algorithm.
